Vragenlijst thesis evaluatie

### 2015-08-16 doc/template/chap-5.tex
Samenvatting nog verder verbeterd
Alles

One of the assumptions underlying game theory is that the players of the game are independent and self-interested. This does not necessarily mean that they want to harm other agents or that they only care about themselves. 
%utility function meaning 
Instead it means that each agent has preferences about which states of the world he likes. These preferences are mapped to natural numbers and are called the utility functions. The numbers are interpreted as a mathematical measure that tells how much an agent likes or dislikes the outcome of the game. Outcomes are the result of a specific combination of a player's strategy. Each combination of a player's strategy is an outcome of the game. \\% The utility function can be ordinal or cardinal. In case of a ordinal utility function only relative rankings are important. The utility function can also be cardinal which is important for games with mixed strategies. \\

%Cooperative and non cooperative games
%-Remark- Stuk herschrijven-- Payoff introduceren -- zin van wikipedia
%Binnen de speltheorie is een 'spel' een (model van) interactie tussen spelers die individueel beslissingen nemen. De spelers ontvangen een beloning (pay-off) die afhangt van hun eigen beslissingen maar ook van die van de anderen. De volgende typen kunnen worden onderscheiden.
%
%Coöperatieve spellen en niet-coöperatieve spellen[bewerken]
%Het verschil tussen deze twee typen spellen is dat bij coöperatieve spellen (cooperative game theory) bindende afspraken tussen de spelers gemaakt kunnen worden terwijl dit niet mogelijk is bij de niet-coöperatieve spellen (non-cooperative game theory).
%The choices of each agent produce an outcome, with respect to the utilities of the agent. A pay-off of an agent is the 
\todo{utility, pay-off, outcomes}
Games can be divided into two types of games: cooperative or non-cooperative. In a non cooperative game, the basic modelling unit is the group of agents. 
Two or more agents want to maximise their utility and their actions can be affected by other agents' utilities. In the individualistic approach the basic modelling is only one agent. This type of game is referred to as \textit{cooperative game theory}.

@@

If there is no feedback for either player, he will play in the same manner against every opponent. The strategy is then called \textit{non-adaptive} because the playing strategy is not dependent on the opponent's movements. An interesting subclass of the non-adaptive strategies are the Renewal strategies where the time intervals between two consecutive moves are generated by a renewal process. The length between two consecutive moves are independent and identically distributed random variables chosen from a density formula \textit{f}. This renewal process means that the strategies are renewed after each move: the interval until the next move only depends on the current move time and not on previous history. \todo{check comment in latex}
%-Remark- Vorige zin is gecopypaste van FlipIt
An example of such renewal strategy is the periodic strategy where the time between two consecutive moves of the players is a fixed interval. An exponential strategy is a renewal strategy in which the interval between two consecutive moves is exponentially distributed. \\

>>>>

One of the assumptions underlying game theory is that the players of the game are independent and self-interested. %This does not necessarily mean that they want to harm other agents or that they only care about themselves. 
%utility function meaning 
This means they do not actively seek to harm other agents. Instead, each agent has preferences about the state of the world. These preferences are mapped to natural numbers in a function called the utility function. These utility values are a mathematical measure indicating how much an agent likes or dislikes the outcome of the game. Outcomes are the result of a specific combination of a player's strategy. Each combination of a player's strategy is an outcome of the game. A rational, independent and self-interested player prefers outcomes with a higher personal utility value, with no regard for the outcome of other players.\\

% The utility function can be ordinal or cardinal. In case of a ordinal utility function only relative rankings are important. The utility function can also be cardinal which is important for games with mixed strategies. \\

%Cooperative and non cooperative games
%-Remark- Stuk herschrijven-- Payoff introduceren -- zin van wikipedia
%Binnen de speltheorie is een 'spel' een (model van) interactie tussen spelers die individueel beslissingen nemen. De spelers ontvangen een beloning (pay-off) die afhangt van hun eigen beslissingen maar ook van die van de anderen. De volgende typen kunnen worden onderscheiden.
%
%Coöperatieve spellen en niet-coöperatieve spellen[bewerken]
%Het verschil tussen deze twee typen spellen is dat bij coöperatieve spellen (cooperative game theory) bindende afspraken tussen de spelers gemaakt kunnen worden terwijl dit niet mogelijk is bij de niet-coöperatieve spellen (non-cooperative game theory).
%The choices of each agent produce an outcome, with respect to the utilities of the agent. A pay-off of an agent is the

Games can be divided into two types of games: cooperative or non-cooperative. In \textit{non cooperative games}, the basic modelling unit is the group of agents. 
Two or more agents want to maximise their utility and their actions can be affected by other agents' utilities. In the individualistic approach the basic modelling is only one agent. This type of game is referred to as \textit{cooperative games}.

@@

If there is no feedback for either player, he will play in the same manner against every opponent. The strategy is then called \textit{non-adaptive} because the playing strategy is not dependent on the opponent's movements. An interesting subclass of the non-adaptive strategies are the Renewal strategies where the time intervals between two consecutive moves are generated by a renewal process. The length between two consecutive moves are independent and identically distributed random variables chosen from a density formula \textit{f}. This renewal process means that the strategies are renewed after each move: the interval until the next move only depends on the current move time and not on previous history. 
%-Remark- Vorige zin is gecopypaste van FlipIt
An example of such renewal strategy is the periodic strategy where the time between two consecutive moves of the players is a fixed interval. An exponential strategy is a renewal strategy in which the interval between two consecutive moves is exponentially distributed. \\

### 2015-08-16 doc/template/chap-9.tex
Alles

\section{Introduction}
\label{Ch2:Intro}
The FlipIt game with propagation delay considers a game where the moves of the attacker are not instantaneous. This corresponds to APT's which use malware - viruses, worms or trojan horses - to perform attacks, as the malware needs time to propagate. A virus, for example, can be dropped on a network but it only compromises the whole network if every node in the network is infected. So there is a certain delay between the moment of attack and the moment the attacker has control over the resource, called the propagation delay. The basic FlipIt game does not take this propagation delay into account. This chapter explains how the FlipIt game with propagation delay can be modelled. Section \ref{ch2:diffFlip} explains the difference between a basic FlipIt game and a FlipIt game with propagation delay. The last section \ref{ch2:periodicvirus} derives a formula to calculate the benefit for a FlipIt game with propagation delay. In the next Chapter, this benefit formula will be used to determine what the best defence and attack strategies are.
%Section \todo{nash} presents the Nash equilibrium for the benefit formula. The chapter concludes in section \todo{section toevoegen} with lessons learned en practical implications for cyber security.
%
%\section{FlipIt game with virus propagation}
%
%Welke aanpassing en motivatie: attacker 1 speler die elke keer virus probeert te droppen. Defender die elke keer patcht als er virus gedropt wordt. Elke patch voldoet aan de meest recente patch dus alle virussen die ook gedropt zijn worden weggedaan. Voor future work: patch systeem aanpassen. 
%
%The actions of a defender: the actions of an attacker:

@@

The network is composed of multiple nodes. The defender can choose to flip one node, a subset of nodes or all nodes of the network. In this paper, the defender flips all the nodes in the network every time he plays. This action can be extended to flipping only a subset of the nodes or a single node of the network. This extension has already been investigated in the context of placing anti-virus systems in a graph. This problem is known to be NP-hard. \cite{NPhard}. \\ \todo{niet volledig juist, defender systemen geplaatst en dan gezien hoe ver de worm kan geraken}

@@

In this paper we assume that both moves of the attacker and the defender are stealthy. Our motivation for a defender with stealthy moves is that it can not be assumed that every attacker can receive feedback from an APT. Some APT's can be resident on a isolated network (e.g. Stuxnetworm), meaning that there is no way to connect to the internet. \\

@@

Even when the APT is launched to steal sensitive information the defender can set-up a honey pot, making the attacker believe that he is stealing sensitive information, but the information in the honeypot is all fake. Honeypots emulate services or creates multiple instances of real operating systems and can pretend to have sensitive information.  Honeypots do not detect all malicious attacks so the attacker can stay unnoticed and can still provide information as feedback. Furthermore, the case where the moves of the defender are non stealthy has already been investigated by Laszka \citep{MitigationNonTargeted}.

@@

\textbf{Periodic strategy}\\
The choice of periodic strategy is motivated by the assumption that in most organisations, the defence strategy is to periodically defend the network.  This is true for the example of patching. Companies like Microsoft and Google release their security patches at fixed intervals, so companies will apply these patches as they come out. Microsoft security bulletins are released on the Second Tuesday of each month. \cite{MicroPatch} \\

@@

\begin{figure}[hbtp]
\centering
\includegraphics[scale=0.7]{Images/Delayuitgelegd.pdf}
\caption{The first game is the basic FlipIt game. The second is the FlipIt game with a delay. During the first flip of the attacker, the defender moves after the delay, causing the attacker to get control over the resource. During the second flip of the attacker, the defender flips at time t+d, causing the defender to take control over the resource before the attacker. During the third and final flip of the attacker, the defender flips during time t+d, causing the attacker to never gain control over the resource. }
\label{dt}
\end{figure}

@@

\subsection*{Case 0: $\delta_{D} \leq d$ (The defender plays with a period smaller than the delay}

@@

However, if the delay is larger than or equal to the period of the defender, the attacker will never gain control over the resource. The defender will always flip before the delay is over. This situation is represented in figure \ref{langeredelay}, where the attacker plays with an interval of 5 time units, the delay is equal to 6 time units and the defender plays with an interval of 4 time units. \\
For the attacker it does not matter if he plays faster or slower than the defender. The benefit of the defender and the attacker are as follows:
\begin{equation}\label{case0def}
\beta_{D}(\delta_{D},\delta_{A})=1-\dfrac{k_{D}}{\delta_{D}}
\end{equation}
\begin{equation}\label{case0att}
\beta_{A}(\delta_{D},\delta_{A})=-\dfrac{k_{A}}{\delta_{A}}
\end{equation}
%-Remark- hier staat iets dat niet juist is.
%From this we can conclude that it only makes sense to calculate the formulas for the cases where $d$ is smaller than $\delta_{D}$. We can already conclude that it is of no use for the attacker to play when the delay is bigger than $\delta_{D}$.

@@

For the rest of the cases we assume that $\delta_{D} \geq d$.

@@

 The benefit of the defender is then as follows:
\begin{equation}
\beta_{D}(\delta_{D},\delta_{A}) = 1 - \dfrac { (\delta_{D}-d) ^{2}} {2 \cdot \delta_{D}  \delta_{A}} - \dfrac{k_{D}}{ \delta_{D}}
\label{Benfcase1:defender}
\end{equation}
~~\\
For this case where $d$=0, we this equals the formula of the original FlipIt game \citep{FlipIt} [p675].\\
For the cornercase with case 0, when $d=\delta_{D}$, the benefit of the defender and the attacker result in the same benefits as in case 0.

@@

This however results in an overestimation. 
%By simulation the game, it can be notices that the closer $\delta_{A}/\delta_{D}$ is equal to one, the better the approximation. If $\delta_{A}/\delta_{D} = 1$ the result is correct. 
The reason this formula overestimates the benefit of the attacker is that it assumes that the defender is always in control during the delay. However, if the attacker was in control in the previous interval, then he continuous to be in control during the period of the delay, see figure \ref{fig:case2}. This means that the average benefit formulas for this case cannot be derived from one interval only; what happens in the previous interval must be taken into account. \\

We know that the defender's moves are instantaneous. Therefore, it is easier to calculate the benefit of the defender. Because the defender moves slower than the attacker we know that if the defender moves during the interval of the attacker, he only moves once within this interval.

@@

For the case with a delay we consider two cases, Case a and Case b, depending on whether the delay is shorter or longer than the difference between the attacker's and the defender's period.  \\

@@

It is crucial that $ \delta_{D}$ is at least as large as $d + \delta_{A}$. If not, the defender can move during the delay in the interval following the interval in which the defender already moved. This would result in an overlap between the average gain of $\dfrac{\delta_{A}}{2} +d$ and the delay. The above benefit formula would then include to much gain for the defender: the potential overlap during the delay would be counted twice. See figure \ref{countedtwice}\\

\begin{figure}[hbtp]
\centering
\caption{Cases where the delay would be counted twice.}
\includegraphics[scale=0.7]{Images/flipcase2nb.pdf}
\label{countedtwice}
\end{figure}

@@

The benefit formulas of case 1 and case 2.b result in the same formula. Consequently, these two formulas can be brought together by adapting the border conditions. Figure \ref{Borders} shows all the cases and it can be seen that the border condition of case 1 and case 2.b can be brought together. This results in the following figure \ref{endborders}. \\

>>>>

%\section{Introduction}
%\label{Ch2:Intro}
The FlipIt game with propagation delay considers a game where the moves of the attacker are not instantaneous. This corresponds to APT's which use malware - viruses, worms or trojan horses - to perform attacks, as the malware needs time to propagate. A virus, for example, can be dropped on a network but it only compromises the whole network if every node in the network is infected. So there is a certain delay between the moment of attack and the moment the attacker has control over the resource, called the propagation delay. The basic FlipIt game does not take this propagation delay into account. This chapter explains how the FlipIt game with propagation delay can be modelled. Section \ref{ch2:diffFlip} explains the difference between a basic FlipIt game and a FlipIt game with propagation delay. The last section \ref{ch2:periodicvirus} derives a formula to calculate the benefit for a FlipIt game with propagation delay. In the next Chapter, this benefit formula will be used to determine what the best defence and attack strategies are.
%Section \todo{nash} presents the Nash equilibrium for the benefit formula. The chapter concludes in section \todo{section toevoegen} with lessons learned en practical implications for cyber security.
%
%\section{FlipIt game with virus propagation}
%
%Welke aanpassing en motivatie: attacker 1 speler die elke keer virus probeert te droppen. Defender die elke keer patcht als er virus gedropt wordt. Elke patch voldoet aan de meest recente patch dus alle virussen die ook gedropt zijn worden weggedaan. Voor future work: patch systeem aanpassen. 
%
%The actions of a defender: the actions of an attacker:

@@

The network is composed of multiple nodes. The defender can choose to flip one node, a subset of nodes or all nodes of the network. The analyses in this paper is based on the assumption that the defender flips all the nodes in the network every time he plays. This action can be extended to flipping only a subset of the nodes or a single node of the network. This extension has already been investigated in the context of installing anti-virus systems on specific nodes in a graph and investigating how this affects the spreading of malware. 
In chapter \ref{chapter4: Worm propagation}, when discussing future work, the use of a matrix model is discussed to determine the probability that a node is infected by malware.  \\

@@

In this paper we assume that both moves of the attacker and the defender are stealthy. Our motivation for a defender with stealthy moves is that it can not be assumed that every attacker can receive feedback from an APT. Some APT's can be resident on a isolated network (e.g. the Stuxnetworm), meaning that there is no way to connect to the Internet. \\

@@

Even when the APT is launched to steal sensitive information the defender can set-up a honey pot, making the attacker believe that he is stealing sensitive information, but the information in the honeypot is all fake. Honeypots emulate services or creates multiple instances of real operating systems and can pretend to have sensitive information.  The case where the moves of the defender are non stealthy has already been investigated by Laszka \citep{MitigationNonTargeted}.

@@

\textbf{Periodic strategy}\\
The choice of periodic strategy is motivated by the assumption that in most organisations, the defence strategy is to periodically defend the network.  This is true for the example of patching. Companies like Microsoft and Google release their security patches at fixed intervals, so companies will apply these patches as they come out. Microsoft security bulletins are released, for example, on the Second Tuesday of each month. \cite{MicroPatch} \\

@@

\begin{figure}[hbtp]
\centering
\includegraphics[scale=0.7]{Images/Delayuitgelegd.pdf}
\caption{The first game is the basic FlipIt game. The second is the FlipIt game with a delay. During the first flip of the attacker, the defender moves after the delay (point 2), resulting in the attacker gaining control over the resource, until the defender flips. During the second flip of the attacker, the defender flips at time t+d, causing the defender to retain control over the resource before the attacker. During the third and final flip of the attacker, the defender flips during time t+d, causing the attacker to never gain control over the resource. }
\label{dt}
\end{figure}

@@

\subsection*{Case 0: $\delta_{D} \leq d$ (The defender plays with a period smaller than the delay)}

@@

However, if the delay is larger than or equal to the period of the defender, the attacker will never gain control over the resource. The defender will always flip before the delay is over. This situation is represented in figure \ref{langeredelay}, where the attacker plays with an interval of 5 time units, the delay is equal to 6 time units and the defender plays with an interval of 4 time units. \\
When $ \delta_{D} \leq d$, it does not matter for the attacker whether he plays faster or slower than the defender. For this case the benefit of the defender and the attacker are as follows:
\begin{equation}\label{case0def}
\beta_{D}(\delta_{D},\delta_{A})=1-\dfrac{k_{D}}{\delta_{D}}
\end{equation}
\begin{equation}\label{case0att}
\beta_{A}(\delta_{D},\delta_{A})=-\dfrac{k_{A}}{\delta_{A}}
\end{equation}
%-Remark- hier staat iets dat niet juist is.
%From this we can conclude that it only makes sense to calculate the formulas for the cases where $d$ is smaller than $\delta_{D}$. We can already conclude that it is of no use for the attacker to play when the delay is bigger than $\delta_{D}$.

@@

For the remainder of the cases we assume that $\delta_{D} \geq d$.

@@

 The benefit of the defender is then as follows:
\begin{equation}
\beta_{D}(\delta_{D},\delta_{A}) = 1 - \dfrac { (\delta_{D}-d) ^{2}} {2 \cdot \delta_{D}  \delta_{A}} - \dfrac{k_{D}}{ \delta_{D}}
\label{Benfcase1:defender}
\end{equation}
~~\\
For this case where $d$=0, this equals the formula of the original FlipIt game \citep{FlipIt} [p675].\\
For the border case, when $d=\delta_{D}$, the benefit of the defender and the attacker result in the same benefits as in case 0.

@@

This however results in an overestimation. 
%By simulation the game, it can be notices that the closer $\delta_{A}/\delta_{D}$ is equal to one, the better the approximation. If $\delta_{A}/\delta_{D} = 1$ the result is correct. 
The reason this formula overestimates the benefit of the attacker is that it assumes that the defender is always in control during the delay. However, if the attacker was in control in the previous interval, then he continues to be in control during the period of the delay, see figure \ref{fig:case2} situation 5 to 7. This means that the average benefit formulas for this case cannot be derived from one interval only; what happens in the previous interval must be taken into account. \\

We know that the defender's moves are instantaneous. Therefore, it is easier to calculate the benefit of the defender. Because the defender moves more slowly than the attacker we know that if the defender moves during the interval of the attacker, he only moves once within this interval.

@@

For the remainder of the analyses we consider two cases, Case a and Case b, depending on whether the delay is shorter or longer than the difference between the attacker's and the defender's period.  \\

@@

It is crucial that $ \delta_{D}$ is at least as large as $d + \delta_{A}$. If not, the defender can move during the delay in the interval following the interval in which the defender already moved. This would result in an overlap between the average gain of $\dfrac{\delta_{A}}{2} +d$ and the delay. The above benefit formula would then include too much gain for the defender: the potential overlap during the delay would be counted twice. See figure \ref{countedtwice}\\

\begin{figure}[hbtp]
\centering
\includegraphics[scale=0.7]{Images/flipcase2nb.pdf}
\caption{Cases where the delay would be counted twice.}
\label{countedtwice}
\end{figure}

@@

We calculated the benefit for three cases: when $\delta_{D}$ is smaller than $d$. When $\delta_{A}$ is smaller than $\delta_{A}$ and when it is larger than $\delta_{A}$.
The benefit formulas of case 1 and case 2.b result in the same formula. Consequently, these two formulas can be brought together by adapting the border conditions. Figure \ref{Borders} shows all the cases and it can be seen that the border condition of case 1 and case 2.b can be brought together. This results in the following figure \ref{endborders}. \\
